---
title: "Introduction to GeneOntologyViz"
author: "jonghyeonseol"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to GeneOntologyViz}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

## Overview

**GeneOntologyViz** is a comprehensive R package for visualizing PANTHER GO enrichment analysis results. It provides multiple visualization types including bar plots, dot plots, heatmaps, and network graphs to help you explore and communicate your Gene Ontology enrichment findings.

## Installation

```{r eval=FALSE}
# Install from GitHub
# devtools::install_github("jonghyeonseol/Geneontology_Analysis")

# Or install locally
# devtools::install("path/to/GeneOntologyViz")
```

## Quick Start

### Load the Package

```{r eval=FALSE}
library(GeneOntologyViz)
```

### Run Complete Pipeline

The simplest way to use GeneOntologyViz is to run the complete pipeline:

```{r eval=FALSE}
# Process all files in Input directory
run_go_pipeline()

# Custom settings
run_go_pipeline(
  input_dir = "my_data",
  output_dir = "my_results",
  create_heatmap_plots = TRUE,
  create_network_plots = TRUE
)
```

## Individual Visualizations

### Reading Data

First, read your PANTHER GO enrichment file:

```{r eval=FALSE}
data <- read_go_results("Input/Dataset1_GOBP_Up.txt")
head(data)
```

### Bar Plot

Create a bar plot showing gene count per GO term:

```{r eval=FALSE}
barplot <- create_barplot(data, "GO Enrichment - GOBP Up-regulated")
print(barplot)

# Save plot
ggsave("barplot.png", barplot, width = 12, height = 8, dpi = 300)
```

### Dot Plot

Create a dot plot showing gene ratio:

```{r eval=FALSE}
dotplot <- create_dotplot(data, "GO Enrichment - GOBP Up-regulated")
print(dotplot)
```

### Heatmap

Compare multiple datasets with a heatmap:

```{r eval=FALSE}
# Read multiple datasets
data1 <- read_go_results("Input/Dataset1_GOBP_Up.txt")
data2 <- read_go_results("Input/Dataset2_GOBP_Up.txt")

# Create named list
data_list <- list("Dataset1" = data1, "Dataset2" = data2)

# Generate heatmap
heatmap <- create_heatmap(
  data_list,
  value_type = "pvalue",  # or "fold_enrichment", "count"
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  title = "GO Enrichment Comparison"
)
print(heatmap)
```

### Network Graph

Visualize relationships between GO terms:

```{r eval=FALSE}
network <- create_network(
  data,
  similarity_threshold = 0.3,
  layout = "fr",  # or "circle", "star", "kk"
  node_size_by = "count",
  title = "GO Term Network"
)
print(network)
```

## Configuration

### Viewing Current Configuration

```{r eval=FALSE}
# Print all configuration settings
print_config()

# Get specific value
get_config("fold_enrichment_threshold")
```

### Customizing Settings

```{r eval=FALSE}
# Change fold enrichment threshold
set_config("fold_enrichment_threshold", 15)

# Change number of top terms to display
set_config("top_n_terms", 30)

# Change plot colors
set_config("color_low", "green")
set_config("color_high", "purple")

# Change output format
set_config("output_format", "pdf")
set_config("output_dpi", 600)
```

## Input File Format

GeneOntologyViz expects PANTHER Overrepresentation Test output files with the following structure:

- Line 6: Analysis Type
- Line 7: Annotation Version
- Line 8: Analyzed List with gene count
- Line 9: Reference List
- Line 10: Test Type (FISHER)
- Line 11: Correction (FDR)
- Line 12: Header row
- Line 13+: Data rows

File naming convention: `Dataset{N}_{GOCAT}_{Direction}.txt`

Where:
- N: Dataset number (1, 2, etc.)
- GOCAT: GO Category (GOBP, GOCC, GOMF)
- Direction: Regulation (Up or Down)

## Advanced Usage

### Custom Pipeline

```{r eval=FALSE}
# Custom processing
input_files <- list.files("Input", pattern = "\\.txt$", full.names = TRUE)

results <- list()
for (file in input_files) {
  data <- read_go_results(file)

  if (!is.null(data)) {
    # Create custom visualizations
    plot1 <- create_barplot(data, basename(file))
    plot2 <- create_dotplot(data, basename(file))

    # Store results
    results[[basename(file)]] <- data
  }
}

# Create comparison heatmap
heatmap <- create_heatmap(results, value_type = "fold_enrichment")
```

### Filtering and Subsetting

```{r eval=FALSE}
# Read data
data <- read_go_results("Input/Dataset1_GOBP_Up.txt")

# Filter by p-value
significant_terms <- data[data$P.value < 0.01, ]

# Select top 10 by fold enrichment
top_enriched <- data[order(-data$Fold_Enrichment), ][1:10, ]

# Create visualization
plot <- create_barplot(top_enriched, "Top 10 Enriched Terms")
```

## Tips and Best Practices

1. **File Organization**: Keep input files in a dedicated directory
2. **Naming Convention**: Follow the recommended naming pattern for automatic grouping
3. **Quality Control**: Check input files for proper format before processing
4. **Visualization Selection**:
   - Use bar plots for absolute gene counts
   - Use dot plots for proportional comparisons
   - Use heatmaps for multi-dataset comparisons
   - Use network graphs for exploring term relationships
5. **Parameter Tuning**:
   - Adjust `fold_enrichment_threshold` based on your data
   - Modify `top_n_terms` to show more or fewer terms
   - Change `similarity_threshold` for network density

## Troubleshooting

### No significant results found

- Check fold enrichment threshold (default: ≥10)
- Verify input file format
- Ensure file has enough data lines

### Network plot too dense

- Increase `similarity_threshold` (default: 0.3)
- Reduce `top_n_terms` to show fewer nodes

### Heatmap not generated

- Ensure at least 2 datasets are provided
- Check that datasets have common GO categories

## Getting Help

For more information:
- GitHub Issues: https://github.com/jonghyeonseol/Geneontology_Analysis/issues
- Package Documentation: `?GeneOntologyViz`
- Function Help: `?run_go_pipeline`, `?create_barplot`, etc.

## Citation

If you use GeneOntologyViz in your research, please cite:

```
GeneOntologyViz: Gene Ontology Enrichment Analysis Visualization
jonghyeonseol (2025)
https://github.com/jonghyeonseol/Geneontology_Analysis
```
